---
title: "Hierarchical model: SBC"
subtitle: Prior SBC can't recommend model parameterisation  
author: Teemu Säilynoja
date: 2024-11-08
date-modified: today
format:
  html:
    toc: true
    number-sections: true
    code-copy: hover
    code-tools: true
    code-fold: true
    self-contained-math: true
    standalone: true
    code-overflow: wrap
    message: false
    warning: false
    fig-format: svg
  pdf:
    toc: true
    number-sections: true
    echo: false
    message: false
    warning: false
    fig-format: pdf
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
    include-before-body:
      text: |
        \RecustomVerbatimEnvironment{verbatim}{Verbatim}{
          showspaces = false,
          showtabs = false,
          breaksymbolleft={},
          breaklines
        }
---

```{r}
library("bayesplot")
library("ggplot2")
library("ggforce")
library("SBC")
library("scales")
source("../R/modified_sbc_plots.R")
knitr::opts_knit$set(root.dir = "..")

bayesplot_theme_set(
  theme_default(base_size = 28, base_family = "sans") +
  theme(axis.text.y = element_blank())
)
theme_set(
  theme_default(base_size = 28, base_family = "sans") +
  theme(axis.text.y = element_blank())
)

colors = c("#7fc97f", "#666666", "#ffff99", "#f0027f", "#bf5b17", "#386cb0")

bayesplot::color_scheme_set(colors)
options(ggplot2.discrete.fill = colors)
options(ggplot2.discrete.color = colors)

names_in_paper <- list(
  "mu" = "$\\mu_0$",
  "tau" = "$\\tau$",
  "sigma" = "$\\sigma$",
  "loglik" = "log-likelihood"
  )
  
label_for_paper <- function(labels, ...) {
  ggforce::label_tex(
    dplyr::mutate_all(
      labels, \(col) names_in_paper[col]
    ),
    ...
  )
}
```

```{r}
res <- list(
  prior_cp = readRDS("./hierarchical-model/sbc_results/prior_sbc_cp_niter_500_seed_8769.rds")$result,
  prior_ncp = readRDS("./hierarchical-model/sbc_results/prior_sbc_ncp_niter_500_seed_8769.rds")$result,
  posterior_cp_weak = readRDS("./hierarchical-model/sbc_results/posterior_sbc-cp-weak_likelihood-niter-500-seed-2024.rds")$result,
  posterior_ncp_weak = readRDS("./hierarchical-model/sbc_results/posterior_sbc-ncp-weak_likelihood-niter-500-seed-2024.rds")$result,
  posterior_cp_strong = readRDS("./hierarchical-model/sbc_results/posterior_sbc-cp-strong_likelihood-niter-500-seed-2024.rds")$result,
  posterior_ncp_strong = readRDS("./hierarchical-model/sbc_results/posterior_sbc-ncp-strong_likelihood-niter-500-seed-2024.rds")$result
)
```

Below, we have 50 groups of 5 observations with

\begin{align*}
\mu_0 &\sim \mathcal N(0,1), \\
\tau, \sigma &\sim \mathcal N^+(0,1), \\
\mu_j &\sim \mathcal N( \mu_0, \tau), \\ 
y_{j,i} &\sim \mathcal N(\mu_j, \sigma).\\
\end{align*}


\newpage

## Prior SBC

```{r}
#| layout-ncol: 2
#| label: hierarchical-model-priorSBC
#| fig-height: 8
#| fig-width: 8
#| fig-cap: 
#|   - "Centered"
#|   - "Non-centered"
with(subset(
  res$prior_cp$stats,
  !startsWith(x = res$prior_cp$stats$variable, prefix = "theta")
), {
  modified_ppc_pit_ecdf_grouped(
    pit = as.vector(rank / max_rank),
    group = sapply(variable, \(v) names_in_paper[[v]]),
    facet_labeller = "label_tex",
    plot_diff = TRUE
  ) +
  #ggtitle("Prior SBC - CP") +
  scale_x_continuous(breaks = breaks_pretty(n = 3)) +
  scale_y_continuous(breaks = breaks_pretty(n = 3))
})

with(subset(
  res$prior_ncp$stats,
  !startsWith(x = res$prior_cp$stats$variable, prefix = "theta")
), {
  modified_ppc_pit_ecdf_grouped(
    pit = as.vector(rank / max_rank),
    group = sapply(variable, \(v) names_in_paper[[v]]),
    facet_labeller = label_tex,
    plot_diff = TRUE
  ) +
  #ggtitle("Prior SBC - NCP") +
  scale_x_continuous(breaks = breaks_pretty(n = 3))
})
```


Prior SBC for both CP and NCP looks good. No noticeable miscalibration.


### Centered parametrisation

::: {.callout-note appearance="minimal" icon="false" collapse="true" title="SBC diagnostic messages"}

```{r}
#| code-fold: false
res$prior_cp |> summary()
```
::: 


### Non-centered parametrisation
::: {.callout-note appearance="minimal" icon="false" collapse="true" title="SBC diagnostic messages"}

```{r}
#| code-fold: false
summary(res$prior_ncp)
```
::: 

## Posterior SBC

### Weak likelihood

Generate data using $\mu_0 = 0$, $\tau_w$ = `r round(qnorm(.5 + .5 * .05, 0, 1),2)` (the 5\% quantile), and $\sigma$ = `r round(qnorm(.5 + .5 * .95, 0, 1),2)` (the 95\% quantile).


```{r}
#| layout-ncol: 2
#| label: hierarchical-model-posteriorSBC-weak_likelihood
#| fig-height: 8
#| fig-width: 8
#| fig-cap: 
#|   - "Centered"
#|   - "Non-centered"
with(subset(
  res$posterior_cp_weak$stats,
  !startsWith(x = res$prior_cp$stats$variable, prefix = "theta")
), {
  modified_ppc_pit_ecdf_grouped(
    pit = as.vector(rank / max_rank),
    group = sapply(variable, \(v) names_in_paper[[v]]),
    facet_labeller = label_tex,
    plot_diff = TRUE
  ) +
    #ggtitle("Posterior SBC - CP: Weak likelihood") +
    scale_x_continuous(breaks = breaks_pretty(n = 3)) + scale_y_continuous(breaks = breaks_pretty(n = 3))
})
with(subset(
  res$posterior_ncp_weak$stats,
  !startsWith(x = res$prior_cp$stats$variable, prefix = "theta")
), {
  modified_ppc_pit_ecdf_grouped(
    pit = as.vector(rank / max_rank),
    group = sapply(variable, \(v) names_in_paper[[v]]),
    facet_labeller = label_tex,
    plot_diff = TRUE
  ) +
    #ggtitle("Posterior SBC - NCP: Weak likelihood") +
    scale_x_continuous(breaks = breaks_pretty(n = 3)) + scale_y_continuous(breaks = breaks_pretty(n = 3))
})
```

For the centered parameterisation, we see a dip in the PIT-ECDF of $\tau$. The dip is visible also
for the non-centered parameterisation, with also the joint log-likelihood looking suspicious.

::: {.callout-note appearance="minimal" icon="false" collapse="true" title="SBC diagnostic messages"}

```{r}
#| code-fold: false
summary(res$posterior_cp_weak)
```
::: 

::: {.callout-note appearance="minimal" icon="false" collapse="true" title="SBC diagnostic messages"}

```{r}
#| code-fold: false
summary(res$posterior_ncp_weak)
```
::: 


### Strong likelihood

Generate data using $\mu_0 = 0$, $\tau_w$ = `r round(qnorm(.5 + .5 * .95, 0, 1),2)` (the 95\% quantile), and $\sigma$ = `r round(qnorm(.5 + .5 * .05, 0, 1),2)` (the 5\% quantile).

```{r}
#| layout-ncol: 2
#| label: hierarchical-model-posteriorSBC-strong_likelihood
#| fig-height: 8
#| fig-width: 8
#| fig-cap: 
#|   - "Centered"
#|   - "Non-centered"
with(subset(
  res$posterior_cp_strong$stats,
  !startsWith(x = res$prior_cp$stats$variable, prefix = "theta")
), {
  modified_ppc_pit_ecdf_grouped(
    pit = as.vector(rank / max_rank),
    group = sapply(variable, \(v) names_in_paper[[v]]),
    facet_labeller = label_tex,
    plot_diff = TRUE
  ) +
    #ggtitle("Posterior SBC - CP: Strong likelihood") +
    scale_x_continuous(breaks = breaks_pretty(n = 3)) + scale_y_continuous(breaks = breaks_pretty(n = 3))
})
with(subset(
  res$posterior_ncp_strong$stats,
  !startsWith(x = res$prior_cp$stats$variable, prefix = "theta")
), {
  print(modified_ppc_pit_ecdf_grouped(
    pit = as.vector(rank / max_rank),
    group = sapply(variable, \(v) names_in_paper[[v]]),
    facet_labeller = label_tex,
    plot_diff = TRUE
  ) +
    #ggtitle("Posterior SBC - NCP: Strong likelihood") +
    scale_x_continuous(breaks = breaks_pretty(n = 3)) + scale_y_continuous(breaks = breaks_pretty(n = 3))
  )
})
```

This looks good.

The centered parameterisation is suitable for the observsation.

Although not visible in the joint log-likelihood, if calibrated estimation of the population level parameters is important, the non-centered parameterisation would not be a good choice.

::: {.callout-note appearance="minimal" icon="false" collapse="true" title="SBC diagnostic messages"}
```{r}
#| code-fold: false
summary(res$posterior_cp_strong)
```
::: 

::: {.callout-note appearance="minimal" icon="false" collapse="true" title="SBC diagnostic messages"}
```{r}
#| code-fold: false
summary(res$posterior_ncp_strong)
```
::: 
