---
title: "Lotka-Volterra predator-prey model: SBC"
subtitle: "Hudson's Bay Company lynx-hare data"
author: "Teemu Säilynoja"
date: 2024-07-03
date-modified: today
format:
  html:
    toc: true
    number-sections: true
    code-copy: hover
    code-tools: true
    code-fold: true
    self-contained-math: true
    standalone: true
    code-overflow: wrap
    message: false
    warning: false
    fig-format: svg
  pdf:
    number-sections: true
    echo: false
    message: false
    warning: false
    fig-format: pdf
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
    include-before-body:
      text: |
        \RecustomVerbatimEnvironment{verbatim}{Verbatim}{
          showspaces = false,
          showtabs = false,
          breaksymbolleft={},
          breaklines
        }
---

In this notebook, we collect the results of the four other SBC notebooks 
handling the Lotka-Volterra model.
 
```{r}
#| output: false
#| code-summary: Imports & setup
#| label: Imports & setup
library("bayesplot")
library("SBC")
library("ggplot2")
library("ggforce")
library("dplyr")
library("extrafont")
knitr::opts_knit$set(root.dir = "..")
source("../R/modified_sbc_plots.R")
colors <- c("#7fc97f", "#666666", "#ffff99", "#f0027f", "#bf5b17", "#386cb0")

bayesplot_theme_set(theme_default(base_size = 20, base_family = "sans"))
bayesplot::color_scheme_set(colors)
theme_set(theme_default(base_size = 20, base_family = "sans"))
options(ggplot2.discrete.fill = colors)
options(ggplot2.discrete.color = colors)

total_run_time <- function(sbc_outputs) {
  sbc_outputs |>
    unlist() |>
    stringr::str_match("Total execution time: ([0-9]+.[0-9]+) seconds.") |>
    as.numeric() |>
    suppressWarnings() |>
    Filter(f = \(n) !is.na(n)) |>
    sum() |>
    as.difftime(units = "secs")
}
```

```{r}
#| code-summary: Load results
#| label: Laad results
res_prior_raw <- readRDS("./lotka-volterra/sbc_results/results-prior_sbc-n_iter_250seed_845.rds")$result

res_prior_pth <- readRDS("./lotka-volterra/sbc_results/results-prior_sbc-pth-n_iter_250seed_845.rds")$result

res_posterior_raw <- readRDS("./lotka-volterra/sbc_results/results-posterior_sbc-n_iter_500_seed_2024.rds")$result

res_posterior_pth <- readRDS("./lotka-volterra/sbc_results/results-posterior_sbc_init-n_iter_500_seed_2024.rds")$result
```

```{r}
#| code-summary: Change variable names
#| label: change-variable-names
names_in_paper <- list(
  "theta[1]" = "$\\alpha$",
  "theta[2]" = "$\\beta$",
  "theta[3]" = "$\\gamma$",
  "theta[4]" = "$\\delta$",
  "z_init[1]" = "$H_0$",
  "z_init[2]" = "$L_0$",
  "sigma[1]" = "$\\sigma_h$",
  "sigma[2]" = "$\\sigma_l$",
  "loglik" = "log-likelihood"
)

label_for_paper <- function(labels, ...) {
  ggforce::label_tex(
    dplyr::mutate_all(
      labels, \(col) names_in_paper[col]
    ),
    ...
  )
}
```

\newpage

## Prior SBC

### First attempt

::: {.callout-note appearance="minimal" icon="false" collapse="true" title="SBC diagnostic messages"}
```{r}
#| code-fold: false
summary(res_prior_raw)
```
:::

According to the SBC diagnostics, roughly one in ten of the prior SBC iterations had some divergent
transitions, or iterations that saturated the maximum treedepth. Additionally, we received warnings
of high $\hat R$ values. All of the iterations had some steps rejected. Some rejections are common
at the early stages of warm-up stage, but shouldn't continue to the sampling stage after warm-up.
Three fits failed completely.

Running `r length(res_prior_raw$fits)` SBC iterations took
`r total_run_time(res_prior_raw$outputs) |> prettyunits::pretty_dt()`.

```{r}
#| label: prior-sbc
#| fig-width: 10
#| fig-height: 5
with(res_prior_raw$stats, {
  print(modified_ppc_pit_ecdf_grouped(
    pit = as.vector(
      rank[variable != "loglik"] / max_rank[variable != "loglik"]
    ),
    group = variable[variable != "loglik"],
    plot_diff = T,
    facet_labeller = label_for_paper
  ) +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank()
    )
  )

  ppc_pit_ecdf(
    pit = as.vector(
      rank[variable == "loglik"] / max_rank[variable == "loglik"]
    ),
    plot_diff = T
  ) +
    scale_x_continuous(breaks = scales::breaks_extended()) +
    scale_y_continuous(breaks = scales::breaks_extended(3))
})  
```

```{r}
#| label: prior-sbc-recovery
#| fig-width: 11
#| fig-height: 2
modified_plot_sim_estimated(
  res_prior_raw$stats,
  variables = names(names_in_paper[names_in_paper != "log-likelihood"]),
  facet_labeller = label_for_paper,
  alpha = 1, nrow = 1) +
  # scale_x_continuous(breaks = scales::breaks_extended(
  #   n = 3,
  #   only.loose = TRUE,
  #   w = c(0.5, 0.03, 0.3, 0.1),
  #   Q = c(0, 0.1 * 1:5, 1:5, 10 * (1:5), 100 * (1:5)))) +
  # scale_y_continuous(breaks = scales::breaks_extended(
  #   n = 3,
  #   only.loose = TRUE,
  #   w = c(0.5, 0.05, 0.4, 0.05),
  #   Q = c(0, 0.1 * 1:5, 1:5, 10 * (1:5), 100 * (1:5))
  # )) +
  labs(x = "Ground truth", y = "Estimate") +
    theme_default(base_size = 18, base_family = "sans") +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank()
    ) +
    NULL
```

From the PIT-ECDF plots, we see that there seems to be some tendency to underestimate the joint
log-likelihood. We see that some iterations had especially bad parameter recovery.

### Using Pathfinder for initialization

::: {.callout-note appearance="minimal" icon="false" collapse="true" title="SBC diagnostic messages"}
```{r}
#| code-fold: false
summary(res_prior_pth)
```
:::

We attempt at improving the inference by initializing the Markov chains with Pathfinder. We don't
anymore have fits that would fail completely, and have in general improved the convergence
diagnostics. Still the inference is not without problems. This time, running
`r length(res_prior_pth$fits)` SBC iterations took
`r total_run_time(res_prior_pth$outputs) |> prettyunits::pretty_dt()`.

```{r}
#| label: prior-sbc-pth
#| fig-width: 10
#| fig-height: 5
with(res_prior_pth$stats, {
  print(modified_ppc_pit_ecdf_grouped(
    pit = as.vector(
      rank[variable != "loglik"] / max_rank[variable != "loglik"]
    ),
    group = variable[variable != "loglik"],
    plot_diff = T,
    facet_labeller = label_for_paper
  ) +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank()
    )
  )

  ppc_pit_ecdf(
    pit = as.vector(
      rank[variable == "loglik"] / max_rank[variable == "loglik"]
    ),
    plot_diff = T
  ) +
    scale_x_continuous(breaks = scales::breaks_extended(2)) +
    scale_y_continuous(breaks = scales::breaks_extended(3))
})
```

```{r}
#| label: prior-sbc-pth-recovery
#| fig-width: 11
#| fig-height: 2
modified_plot_sim_estimated(
  res_prior_pth$stats,
  variables = names(names_in_paper[names_in_paper != "log-likelihood"]),
  facet_labeller = label_for_paper,
  alpha = 1, nrow = 1) +
  labs(x = "Ground truth", y = "Estimate") +
    theme_default(base_size = 18, base_family = "sans") +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank()
    ) +
  NULL
```
The PIT-ECDF plots look very similar, and we still see some cases of bad parameter recovery.

### Conclusion

Even with the Pathfinder initialization, the inference isn't calibrated and we can't trust the model
to in general give us trustworthy inference results.

One could try to improve the priors, or the computational aspects of the model. Perhaps adjusting
the tolerance of the ODE solver, or increasing the adapt delta and max treedepth parameters of the
NUTS algorithm could improve the results.

As the inference is relatively slow, iterating on the model is quite slow.

## Posterior SBC

Next, we condition our analysis on the historical Hudson Bay Company data of lynxes and hares.

### First attempt

::: {.callout-note appearance="minimal" icon="false" collapse="true" title="SBC diagnostic messages"}


```{r}
#| code-fold: false
summary(res_posterior_raw)
```

:::

We still observe issues in the inference. Large $\hat R$ values are more common than with prior SBC,
and we have multiple fits and chains that fail to produce samples.

We do see, that the inference is faster. Running `r length(res_posterior_raw$fits)` SBC iterations
took `r total_run_time(res_posterior_raw$outputs) |> prettyunits::pretty_dt()`.

```{r}
#| label: psoterior-sbc
#| fig-width: 10
#| fig-height: 5
with(res_posterior_raw$stats, {
  print(modified_ppc_pit_ecdf_grouped(
    pit = as.vector(
      rank[variable != "loglik"] / max_rank[variable != "loglik"]
    ),
    group = variable[variable != "loglik"],
    plot_diff = T,
    facet_labeller = label_for_paper
  ) +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank()
    )
  )

  ppc_pit_ecdf(
    pit = as.vector(
      rank[variable == "loglik"] / max_rank[variable == "loglik"]
    ),
    plot_diff = T
  ) +
    scale_x_continuous(breaks = scales::breaks_extended()) +
    scale_y_continuous(breaks = scales::breaks_extended(3))
})

```

```{r}
#| label: posterior-sbc-recovery
#| fig-width: 11
#| fig-height: 2
modified_plot_sim_estimated(
  res_posterior_raw$stats,
  variables = names(names_in_paper[names_in_paper != "log-likelihood"]),
  facet_labeller = label_for_paper,
  alpha = 1, nrow = 1) +
  labs(x = "Ground truth", y = "Estimate") +
    theme_default(base_size = 18, base_family = "sans") +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank()
    ) +
  NULL
```

The PIT-ECDF plots are telling of great issues with the calibration of the posterior inference. The
join log-likelihood is in general over estimated, while especially the estimates of the initial
populations seem to have calibration issues.

### Using Pathfinder for initialization

::: {.callout-note appearance="minimal" icon="false" collapse="true" title="SBC diagnostic messages"}
```{r}
#| code-fold: false
summary(res_posterior_pth)
```

:::

Again, we employ Pathfinder to initialize the Markov chains. The high $\hat R$ values as well as
some individual posterior plots indicate multi-modality. We still have a warning about large $\hat
R$ values, but now even the largest $\hat R$ is only 1.012. Additionally we have no warnings on
divergent transitions or max treedepths. We still have rejections in the sampling, but the maximum
number of rejections was 25 and was likely to take place in the early stages of the warm-up.

We also observe a massive boost in sampling efficiency, as running `r length(res_posterior_pth$fits)` SBC iterations only took `r total_run_time(res_posterior_pth$outputs) |> prettyunits::pretty_dt()`.

```{r}
#| label: psoterior-sbc-pth
#| fig-width: 10
#| fig-height: 5
with(res_posterior_pth$stats, {
  print(modified_ppc_pit_ecdf_grouped(
    pit = as.vector(
      rank[variable != "loglik"] / max_rank[variable != "loglik"]
    ),
    group = variable[variable != "loglik"],
    plot_diff = T,
    facet_labeller = label_for_paper
  ) +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank()
    )
  )

  ppc_pit_ecdf(
    pit = as.vector(
      rank[variable == "loglik"] / max_rank[variable == "loglik"]
    ),
    plot_diff = T
  ) +
    scale_x_continuous(breaks = scales::breaks_extended()) +
    scale_y_continuous(breaks = scales::breaks_extended(3))
})

```

```{r}
#| label: posterior-sbc-pth-recovery
#| fig-width: 11
#| fig-height: 2
modified_plot_sim_estimated(
  res_posterior_pth$stats,
  variables = names(names_in_paper[names_in_paper != "log-likelihood"]),
  facet_labeller = label_for_paper,
  alpha = 1, nrow = 1) +
  labs(x = "Ground truth", y = "Estimate") +
    theme_default(base_size = 18, base_family = "sans") +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank()
    ) +
  NULL
```
All of the PIT-ECDF plots look good, and the parameter recovery is decent.

### Conclusion

After posterior SBC, the inference with Pathfinder initialization looks good enough for us to use
the model.


```{r}
#| label: lv-sbc-comparison
#| fig-height: 5
#| fig-width: 10
with(dplyr::bind_rows(
  cbind(res_prior_pth$stats, data.frame(sbc_type = "Prior SBC")),
  cbind(res_posterior_pth$stats, data.frame(sbc_type = "Posterior SBC"))
), {
  ppc_pit_ecdf_grouped(
    pit = as.vector(
      rank[variable == "loglik"] / max_rank[variable == "loglik"]
    ),
    group = factor(sbc_type[variable == "loglik"], levels = c("Prior SBC", "Posterior SBC")),
    plot_diff = T
  ) +
    scale_x_continuous(breaks = scales::breaks_extended(2)) +
    scale_y_continuous(breaks = scales::breaks_extended(3)) +
    labs(y = "ECDF difference")
})
```