---
title: Posterior SBC on the Lotka-Volterra model
author: Teemu Säilynoja
date: 2024-07-03
date-modified: last-modified
format:
  html:
    toc: true
    number-sections: true
    code-copy: hover
    code-tools: true
    code-fold: true
    self-contained-math: true
    standalone: true
    code-overflow: wrap
    message: false
    warning: false
    fig-format: svg
  pdf:
    number-sections: true
    echo: false
    message: false
    warning: false
    fig-width: 8
    fig-height: 6
    fig-format: pdf
    fig-cap-location: bottom
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
    include-before-body:
      text: |
        \RecustomVerbatimEnvironment{verbatim}{Verbatim}{
          showspaces = false,
          showtabs = false,
          breaksymbolleft={},
          breaklines
        }
params:
  n_sbc_iter: 500
  n_samples: 4000
  iter_warmup: 1000
  seed: 2024
---


In this notebook, we show the results of calibration assessment on the 
[Lotka-Volterra model](https://mc-stan.org/learn-stan/case-studies/lotka-volterra-predator-prey.html) using posterior SBC. We begin by conditioning the model
on the 21 years of observations shown below in @fig-lynx-hare-data.

For collected results of all our experiments, see the notebook named [lotka-volterra-sbc](lotka-volterra-sbc.html).
```{r}
#| label: imports-and-options
#| code-summary: "imports and options"
library(SBC)
library(bayesplot)
library(calculus)
library(cmdstanr)
library(posterior)
library(ggplot2)

options(mc.cores = parallel::detectCores())
colors <- c("#7fc97f", "#666666", "#ffff99", "#f0027f", "#bf5b17", "#386cb0")

bayesplot::color_scheme_set(colors)
options(ggplot2.discrete.fill = colors)
options(ggplot2.discrete.color = colors)
options(ggplot2.binned.color = colors)

theme_set(
  theme_default(base_size = 16, base_family = "sans"))
bayesplot_theme_set(
  theme_default(base_size = 16, base_family = "sans"))
options(mc.cores = parallel::detectCores())
source("../R/modified_sbc_plots.R")

save_stan_fit <- TRUE
stan_fit_dir <- "fits"
cache_dir <- "sbc_results"

names_in_paper <- list(
  "theta[1]" = "$\\alpha$",
  "theta[2]" = "$\\beta$",
  "theta[3]" = "$\\gamma$",
  "theta[4]" = "$\\delta$",
  "z_init[1]" = "$H_0$",
  "z_init[2]" = "$L_0$",
  "sigma[1]" = "$\\sigma_h$",
  "sigma[2]" = "$\\sigma_l$",
  "loglik" = "log-likelihood"
)

label_for_paper <- function(labels, ...) {
  ggforce::label_tex(
    dplyr::mutate_all(
      labels, \(col) names_in_paper[col]
    ),
    ...
  )
}
```

```{r}
#| label: sbc-parameters
thin_by <- 10
n_samples <- params$n_samples
iter_warmup <- params$iter_warmup
n_chains <- 4
n_sbc_iter <- params$n_sbc_iter
seed <- params$seed
set.seed(seed)
```

```{r}
#| label: fig-lynx-hare-data
#| fig-cap: "Historical data of pelts collected by Hudson Bay company."
lynx_hare_df <-
  read.csv("hudson-bay-lynx-hare.csv",
           comment.char = "#")
lynx_hare_df |>
  ggplot(aes(x = Year)) +
  geom_path(aes(y = Lynx,
                colour = "Lynx")) +
  geom_path(aes(y = Hare,
                colour = "Hare")) +
  theme(legend.position = "inside",
        legend.position.inside = c(.9,.95)) +
  labs(y = "Pelts", colour = "") +
  scale_colour_manual(values = c("Hare" = colors[5], "Lynx" = colors[6]))
```

## The Model

```{r}
#| label: The Stan model
#| class-output: stan
#| code-summary: "Load the following Stan model"
model_pred <- cmdstanr::cmdstan_model("models/lotka_volterra_pred.stan")
```

```{r}
#|label: condition posterior
if (file.exists(file.path(stan_fit_dir, "fit0.RDS"))) {
  fit0 <- readRDS(file.path(stan_fit_dir, "fit0.RDS"))
} else {
  fit0 <- model_pred$sample(
    data = list(
      N = nrow(lynx_hare_df) - 1,
      ts = 2:nrow(lynx_hare_df),
      y_init = c(lynx_hare_df$Hare[1], lynx_hare_df$Lynx[1]),
      y = as.matrix(lynx_hare_df[-1, c(3,2)]),
      n_pred = 0
    ),
    seed = seed,
    refresh = 0,
    parallel_chains = n_chains,
    iter_warmup = iter_warmup,
    iter_sampling = n_samples / n_chains
  )
  if (save_stan_fit) {
    fit0$save_object(file.path(stan_fit_dir, "fit0.RDS"))
  }
}
```

We receive no warnings of divergent transitions, or high R hat values, but see strong correlation between many of the parameter posteriors.
```{r}
#| label: fig-posterior-pairs
mcmc_pairs(
  x = fit0$draws(
    variables = c("theta", "sigma")
  ),
  condition = pairs_condition(draws = .5),
  off_diag_args = list(
    alpha = .1,
    size = .5
  )
)
```

## Data conditioned simulation-based calibration
Nex, we run posterior SBC, by augmenting the data with posterior predictive draws and refitting the model to these two datasets.

```{r}
gen_vars <- as_draws_matrix(merge_chains(thin_draws(fit0$draws(
  variables = c("theta", "z_init", "z", "sigma", "loglik")
), n_samples / n_sbc_iter)))
pred_sample <- as_draws_matrix(merge_chains(thin_draws(fit0$draws(
  variables = c("y_init_rep", "y_rep")
), n_samples / n_sbc_iter)))

gen_vars[, "loglik"] <- gen_vars[, "loglik"] +
  sapply(1:nrow(gen_vars), \(k) {
  dlnorm(pred_sample[k, "y_init_rep[1]"], log(gen_vars[k, "z_init[1]"]), gen_vars[k, "sigma[1]"], log = T) +
    dlnorm(pred_sample[k, "y_init_rep[2]"], log(gen_vars[k, "z_init[2]"]), gen_vars[k, "sigma[2]"], log = T) +
    sum(dlnorm(pred_sample[k, paste("y_rep[", 1:20, ",1]", sep = "")], log(gen_vars[k, paste("z[", 1:20, ",1]", sep = "")]), gen_vars[k, "sigma[1]"], log = T)) +
    sum(dlnorm(pred_sample[k, paste("y_rep[", 1:20, ",2]", sep = "")], log(gen_vars[k, paste("z[", 1:20, ",2]", sep = "")]), gen_vars[k, "sigma[2]"], log = T))
})

posterior_datasets <- SBC_datasets(variables = gen_vars[, c(
  "theta[1]",
  "theta[2]",
  "theta[3]",
  "theta[4]",
  "z_init[1]",
  "z_init[2]",
  "sigma[1]",
  "sigma[2]",
  "loglik"
)],
generated = apply(
  pred_sample,
  1,
  \(sample) list(
    N = nrow(lynx_hare_df) - 1,
    ts = 2:nrow(lynx_hare_df),
    y_init2 = sample[1:2],
    y2 = matrix(sample[-c(1, 2)], ncol = 2),
    y_init = c(lynx_hare_df$Hare[1], lynx_hare_df$Lynx[1]),
    y = as.matrix(lynx_hare_df[-1, c(3, 2)])
  )
))
```


```{r}
#| label: The Stan model for SBC
#| class-output: stan
#| code-summary: "Stan model for SBC"
sbc_model <- cmdstanr::cmdstan_model("models/lotka_volterra_postSBC.stan")
```


```{r}
#| label: SBC-backend
stan_backend <- SBC_backend_cmdstan_sample(
  sbc_model, iter_warmup = iter_warmup,
  iter_sampling = n_samples / n_chains,
  chains = n_chains)
```


```{r}
#| label: SBC-results
cache_location <- file.path(
  cache_dir,
  paste(
    "results-posterior_sbc-n_iter_",
    n_sbc_iter,
    "_seed_",
    seed,
    ".rds",
    sep = ""
  )
)

if (file.exists(cache_location)) {
  results_post_sbc <- readRDS(cache_location)$result
} else {
  results_post_sbc <- compute_SBC(
  posterior_datasets,
  stan_backend,
  cache_mode = "results",
  cache_location = file.path(
    cache_dir,
    paste("results-posterior_sbc-n_iter_",
    n_sbc_iter,
    "seed_", seed, sep = "")
  ),
  keep_fits = F,
  thin_ranks = thin_by
)
}
```

::: {.callout-note appearance="minimal" icon="false" collapse="true" title="SBC diagnostic messages"}
```{r}
#| code-fold: true
summary(results_post_sbc)
```
:::

We receive warnings of iterations with high R hat values and one or more divergent transitions. In the calibration assessment in @fig-pit-ecdf we see some apparent calibration issues. This is perhaps not as clear from the parameter recovery plots in @fig-parameter-recovery, but there too we see some outliers and in general large uncertainty. This is
likely due to the often occurring multi-modal posteriors, one of which is later shown in @fig-augmented-pairs.

```{r}
#| label: fig-pit-ecdf
modified_ppc_pit_ecdf_grouped(
  pit = results_post_sbc$stats$rank[results_post_sbc$stats$max_rank == n_samples / thin_by - 1] / results_post_sbc$stats$max_rank[results_post_sbc$stats$max_rank == n_samples / thin_by - 1], group = results_post_sbc$stats$variable[results_post_sbc$stats$max_rank == n_samples / thin_by - 1],
  plot_diff = TRUE,
  nrow = 3,
  facet_labeller = label_for_paper) +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = scales::breaks_extended(3, only.loose = TRUE)) +
  scale_y_continuous(breaks = scales::breaks_extended(3, only.loose = TRUE))
```

```{r}
#| label: fig-parameter-recovery
#| warning: false
modified_plot_sim_estimated(
  results_post_sbc$stats,
  variables = unique(results_post_sbc$stats$variable),
  alpha = 1,
  facet_labeller = label_for_paper,
  nrow = 3) +
  labs(x = "Ground truth", y = "Estimate") +
  scale_x_continuous(breaks = scales::breaks_extended(3, only.loose = TRUE)) +
  scale_y_continuous(breaks = scales::breaks_extended(3, only.loose = TRUE))
```

## A closer look at one augmented posterior
Next, we look at the augmented posterior of the SBC iteration where the worst R hat values were encountered.

```{r}
bad_sim_id <- results_post_sbc$default_diagnostics$sim_id[
  which.max(results_post_sbc$default_diagnostics$max_rhat)]

if (file.exists(file.path(stan_fit_dir, "fit_high_r_hat.RDS"))) {
  fit_high_r_hat <- readRDS(file.path(stan_fit_dir, "fit_high_r_hat.RDS"))
} else {
fit_high_r_hat <- sbc_model$sample(
  data = c(posterior_datasets$generated[[bad_sim_id]], n_pred = 0),
  iter_warmup = iter_warmup,
  iter_sampling = n_samples / n_chains,
  chains = n_chains,
  refresh = 0,
  show_messages = F,
  show_exceptions = F,
  seed = seed
  )
  if (save_stan_fit) {
    fit_high_r_hat$save_object(file.path(stan_fit_dir, "fit_high_r_hat.RDS"))
  }
}
```

Below, in @fig-posterior-draw, we show the parameter values used in generating the posterior predictive draw.
```{r}
#| label: fig-posterior-draw
mcmc_recover_hist(
  x = fit0$draws(colnames(posterior_datasets$variables)),
  true = c(as_draws_matrix(merge_chains(thin_draws(fit0$draws(
  variables = c("theta", "z_init", "sigma", "loglik")
), n_samples / n_sbc_iter)))[bad_sim_id, ]),
  bins = 50
) +
facet_wrap("Parameter", scales = "free", labeller = label_for_paper) +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = scales::breaks_extended(3, only.loose = TRUE)) +
  scale_y_continuous(breaks = scales::breaks_extended(3, only.loose = TRUE))
```

Next, we look at the pairs plot of the parameter posteriors. Below, in @fig-augmented-pairs, we see strong multi-modality as well as correlation between parameter values.

```{r}
#| label: fig-augmented-pairs
#| fig-height: 10
#| fig-width: 10
mcmc_pairs(
  x = fit_high_r_hat$draws(
    variables = c("theta", "sigma")
  ),
  condition = pairs_condition(draws = .5),
  off_diag_args = list(
    alpha = .1,
    size = .5
  )
)
```
```{r}
library(plyr)
dat <- fit_high_r_hat$draws(variables = c("theta[1]", "theta[3]"),
                            format = "df")[, c("theta[1]", "theta[3]")]

id_ls <- dat |> apply(1, \(r) {
  (r[1] - mean(dat$`theta[1]`[dat$`theta[1]` < .8 &
                                dat$`theta[3]` < 1.1])) ^ 2 +
    (r[2] - mean(dat$`theta[3]`[dat$`theta[1]` < .8 &
                                  dat$`theta[3]` < 1.1])) ^ 2
}) |>  which.min()

id_hs <- dat |> apply(1, \(r) {
  (r[1] - mean(dat$`theta[1]`[dat$`theta[1]` >= .8 &
                                dat$`theta[3]` >= 1.1])) ^ 2 +
    (r[2] - mean(dat$`theta[3]`[dat$`theta[1]` >= .8 &
                                  dat$`theta[3]` >= 1.1])) ^ 2
}) |>
  which.min()
```


```{r}
#| label: bimodal-posterior
#| fig-height: 6
#| fig-width: 8
#| fig-cap: Predictive mean and 95\% predictive interval for parameter values from the two modes.
library(latex2exp)
dat_pops <- fit_high_r_hat$draws(variables = c("theta", "z_init", "sigma"),
                                 format = "matrix")
pops_ls <- ode(
  f = function(h, l) {
    c((dat_pops[id_ls, "theta[1]"] - dat_pops[id_ls, "theta[2]"] * l) * h, (-dat_pops[id_ls, "theta[3]"] + dat_pops[id_ls, "theta[4]"] * h) * l)
  },
  var = c(h = dat_pops[id_ls, "z_init[1]"], l = dat_pops[id_ls, "z_init[2]"]),
  times = 1:21
)

pops_hs <- ode(
  f = function(h, l) {
    c((dat_pops[id_hs, "theta[1]"] - dat_pops[id_hs, "theta[2]"] * l) * h, (-dat_pops[id_hs, "theta[3]"] + dat_pops[id_hs, "theta[4]"] * h) * l)
  },
  var = c(h = dat_pops[id_hs, "z_init[1]"], l = dat_pops[id_hs, "z_init[2]"]),
  times = 1:21
)
blue <- "#386cb0"
red <- "#bf5b17"
green <- "#57A773"

scatter_plot <- ggplot(dat) +
  geom_point(
    aes(x = `theta[1]`, y = `theta[3]`),
    colour = "black",
    alpha = .1,
    stroke = 0,
    size = 2
  ) +
  annotate(
    "point",
    x = unlist(dat[id_ls, "theta[1]"]),
    y = unlist(dat[id_ls, "theta[3]"]),
    colour = red,
    size = 5,
    shape = 18
  ) +
  annotate(
    "point",
    x = unlist(dat[id_hs, "theta[1]"]),
    y = unlist(dat[id_hs, "theta[3]"]),
    colour = blue,
    size = 5,
    shape = 18
  ) +
  labs(x = TeX(r"($\alpha$)"),
       y = TeX(r"($\gamma$)")) +
  theme(axis.text = element_blank()) +
NULL

pop_plot <- ggplot(mapping = aes(x = lynx_hare_df$Year)) +
  geom_ribbon(aes(
    ymax = qlnorm(.95, log(pops_hs[, "h"]), dat_pops[id_hs, "sigma[1]"]),
    ymin = qlnorm(.05, log(pops_hs[, "h"]), dat_pops[id_hs, "sigma[1]"])
  ),
  fill = blue,
  alpha = .2) +
  geom_ribbon(aes(
    ymax = qlnorm(.95, log(pops_ls[, "h"]), dat_pops[id_ls, "sigma[1]"]),
    ymin = qlnorm(.05, log(pops_ls[, "h"]), dat_pops[id_ls, "sigma[1]"])
  ),
  fill = red,
  alpha = .2) +
  geom_path(aes(y = qlnorm(.5, log(pops_ls[, "h"]), dat_pops[id_ls, "sigma[1]"])), colour = red) +
  geom_line(aes(y = qlnorm(.5, log(pops_hs[, "h"]), dat_pops[id_hs, "sigma[1]"])), colour = blue) +
  geom_point(aes(y = lynx_hare_df$Hare, shape = "Obs"), colour = "black") +
  geom_point(aes(y = unname(
    c(
      posterior_datasets$generated[[bad_sim_id]]$y_init2[1],
      posterior_datasets$generated[[bad_sim_id]]$y2[, 1]
    )
  ), shape = "Sim"), colour = "gray30") +
  labs(y = "Population", x = "Time", shape = "") +
  theme(
    axis.ticks.y = element_blank(),
    axis.text = element_blank(),
    legend.position = "none",
    legend.box.spacing = unit(-10, "pt"),
    legend.margin = margin(0, 0, 0, 0),
    panel.grid = element_blank()
  )

scatter_plot +
  annotation_custom(grob = ggplotGrob(pop_plot),
                    xmin = 1.3,
                    ymin = 1.1)
```

Lastly, we show the resulting marginal posteriors in relation to the true parameter values.
```{r}
#| label: recover_hist_bad_posterior
#| fig-height: 8
#| fig-width: 10
mcmc_recover_hist(
  x = fit_high_r_hat$draws(colnames(posterior_datasets$variables)),
  true = c(posterior_datasets$variables[bad_sim_id, ]),
  bins = 50
) +
facet_wrap("Parameter", scales = "free", labeller = label_for_paper) +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = scales::breaks_extended(3, only.loose = TRUE)) +
  scale_y_continuous(breaks = scales::breaks_extended(3, only.loose = TRUE))
```


