---
title: Prior SBC on the Lotka-Volterra model
author: Teemu Säilynoja
date: 2024-06-20
date-modified: last-modified
format:
  html:
    toc: true
    number-sections: true
    code-copy: hover
    code-tools: true
    code-fold: true
    self-contained-math: true
    standalone: true
    code-overflow: wrap
    message: false
    warning: false
    fig-format: svg
  pdf:
    number-sections: true
    echo: false
    message: false
    warning: false
    fig-width: 8
    fig-height: 6
    fig-format: pdf
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
    include-before-body:
      text: |
        \RecustomVerbatimEnvironment{verbatim}{Verbatim}{
          showspaces = false,
          showtabs = false,
          breaksymbolleft={},
          breaklines
        }
params:
  n_sbc_iterations: 250
  n_samples: 4000
  iter_warmup: 1000
  seed: 845
---


In this notebook, we show the results of calibration assessment on the 
[Lotka-Volterra model](https://mc-stan.org/learn-stan/case-studies/lotka-volterra-predator-prey.html) using prior SBC.
For collected results of all our experiments, see the notebook named [lotka-volterra-sbc](lotka-volterra-sbc.html).

```{r}
#| label: imports-and-options
#| output: false
#| code-summary: "imports and options"
#| code-fold: true
library(SBC)
library(calculus)
library(cmdstanr)
library(ggplot2)
library(bayesplot)
colors <- c("#7fc97f", "#666666", "#ffff99", "#f0027f", "#bf5b17", "#386cb0")

bayesplot::color_scheme_set(colors)
options(ggplot2.discrete.fill = colors)
options(ggplot2.discrete.color = colors)
options(ggplot2.binned.color = colors)

source("../R/modified_sbc_plots.R")
theme_set(
  theme_default(base_size = 16, base_family = "sans"))
options(mc.cores = parallel::detectCores())

cache_dir <- "sbc_results"

names_in_paper <- list(
  "theta[1]" = "$\\alpha$",
  "theta[2]" = "$\\beta$",
  "theta[3]" = "$\\gamma$",
  "theta[4]" = "$\\delta$",
  "z_init[1]" = "$H_0$",
  "z_init[2]" = "$L_0$",
  "sigma[1]" = "$\\sigma_h$",
  "sigma[2]" = "$\\sigma_l$",
  "loglik" = "log-likelihood"
)

label_for_paper <- function(labels, ...) {
  ggforce::label_tex(
    dplyr::mutate_all(
      labels, \(col) names_in_paper[col]
    ),
    ...
  )
}
```

## Setup
Like the original Hudson Bay data, we simulate data sets of 21 years. We run the SBC on
`r params$n_sbc_iterations` prior predictive samples, each with a total of `r params$n_samples` post
warm-up posterior samples, what will be thinned down by a factor of 10, yielding
`r params$n_samples / 10` posterior draws for calculating the rank of the prior draws and joint log-likelihood.

```{r}
#| label: sbc-parameters
#| echo: false
n_years <- 21
n_sbc_iterations <- params$n_sbc_iterations
n_samples <- params$n_samples
n_chains <- 4
iter_warmup <- params$iter_warmup
thin_by <- 10
seed <- params$seed
set.seed(seed)
```

## The Model
```{r}
#| label: The Stan model
#| class-output: stan
#| code-summary: "Load the following Stan model:"
#| code-fold: true
model <- cmdstanr::cmdstan_model("models/lotka_volterra.stan")
model$print()
```

## Simulation-based calibration

### Prior predictive samples
For implementing the SBC, we use the [SBC R-package](https://hyunjimoon.github.io/SBC/index.html) by Angie H. Moon et al.
First, we define the data generating process, which in our case is generating prior predictive samples.

```{r}
#| label: prior_predictive_generator
prior_predictive_generator_years <- function(n_years){
  # Some prior predictive samples produce negative populations.
  # We reject those, but out of curiosity I'll add a counter to keep
  # track of the number of rejected samples.
  valid_populations <- FALSE
  n_tries <- 0
  while (!valid_populations) {
    n_tries <- n_tries + 1
    # Draw parameter values from the prior:
    alpha <- qnorm(runif(1, pnorm(0, 1, .5), 1), 1, .5)
    beta <- qnorm(runif(1, pnorm(0, .05, .05), 1), .05, .05)
    gamma <- qnorm(runif(1, pnorm(0, 1, .5), 1), 1, .5)
    delta <- qnorm(runif(1, pnorm(0, .05, .05), 1), .05, .05)

    l0 <- rlnorm(1, log(10), 1)
    h0 <- rlnorm(1, log(10), 1)

    # Solve the ODE system to get populations for years 2-21.
    pops <- ode(f = function(h, l) {
      c((alpha - beta * l)*h, (-gamma + delta * h)*l)
      },
                var = c(h = h0, l = l0),
                times = seq(1,n_years, .1))
    if (all(pops > 0)) {
      valid_populations <- TRUE
    }
  }
  
  # With valid populations, draw observation noise from priors and
  # simulate observations.
  sigma_h <- rlnorm(1, -1, 1)
  sigma_l <- rlnorm(1, -1, 1)
  hare_pelts <- rlnorm(n_years, log(pops[seq(1,201,10),1]), sigma_h)
  lynx_pelts <- rlnorm(n_years, log(pops[seq(1,201,10),2]), sigma_l)
  
  # Return the ground truth as `variables` and observations as `generated` 
  list(
    variables = list(
      theta = c(alpha, beta, gamma, delta),
      z_init = c(h0, l0),
      sigma = c(sigma_h, sigma_l),
      loglik = sum(dlnorm(hare_pelts, log(pops[seq(1,201,10), 1]), sigma_h, log = T)) +
        sum(dlnorm(lynx_pelts, log(pops[seq(1,201,10), 2]), sigma_l, log = T))
    ),
    generated = list(
      N = n_years - 1,
      ts = 2:n_years,
      y_init = c(hare_pelts[1], lynx_pelts[1]),
      y = matrix(c(hare_pelts[-1], lynx_pelts[-1]), ncol = 2),
      n_tries = n_tries
    )
  )
}
```

Next, we use the data generating process to define a data generator and construct `r n_sbc_iterations` data sets for SBC.

```{r}
#| label: prior_predictive_datasets
prior_predictive_generator <- SBC_generator_function(
  prior_predictive_generator_years, n_years = n_years)

prior_predictive_datasets <- generate_datasets(
  prior_predictive_generator, n_sims = n_sbc_iterations)

```


```{r}
pr_sample <- sample.int(n_sbc_iterations, 7) |>
  lapply(\(id) {
    data.frame(rbind(
      prior_predictive_datasets$generated[[id]]$y_init,
      prior_predictive_datasets$generated[[id]]$y
      ))   |>
      dplyr::rename(Hare = X1, Lynx = X2) |>
      cbind(data.frame(gen_id = paste("PP", id), Year = 1900:1920))
    }) |>
  dplyr::bind_rows() |> 
  rbind(
    read.csv("hudson-bay-lynx-hare.csv",
           comment.char = "#") |> 
      cbind(data.frame(gen_id = "Observed"))
  )
```


```{r}
#| label: prior-pc-pelt-year
#| fig-cap: Observed data compared to prior predictive draws.
pr_sample |> 
  ggplot() +
  aes(x = Year) +
  geom_path(aes(y = Hare, colour = "Hare")) +
  geom_path(aes(y = Lynx, colour = "Lynx")) +
  facet_wrap(
    ~ gen_id,
    scales = "free_y",
    nrow = 2,
    labeller = as_labeller(\(x) {ifelse(x == "Observed", "Observed", "")})
  ) +
  theme(legend.position = "bottom") +
  labs(y = "Pelts (thousands)", x = "Year", colour = "Species") + scale_color_manual(values = colors[c(5,6)]) +
  scale_x_continuous(breaks = scales::breaks_pretty(2))
```

```{r}
#| label: prior-pc-hare-lynx
#| fig-cap: Trajectories of the population dynamics in the observation and the prior predictive draws from above.
pr_sample |> 
  ggplot() +
  aes(x = Hare, y = Lynx, group = gen_id) +
  geom_path() +
  facet_wrap(vars(gen_id), scales = "free", nrow = 2) +
  theme(legend.position = "bottom") +
  labs(y = "Lynx pelts (thousands)", x = "Hare pelts (thousands)")
```

### Posterior samples
We use `cmdstanr` to obtain posterior draws via MCMC sampling.

```{r}
#| label: cmdstan_backend
stan_backend <- SBC_backend_cmdstan_sample(
  model,
  iter_warmup = iter_warmup,
  iter_sampling = n_samples / n_chains,
  chains = n_chains,
  refresh = 0
)
```

### Run SBC
This is the step requiring a lot of computation. We run MCMC on each of the
predictive samples and store statistics of the results.
```{r}
#| label: run-sbc

cache_location <- file.path(
  cache_dir,
  paste(
    "results-prior_sbc-n_iter_",
    n_sbc_iterations,
    "seed_",
    seed,
    ".rds",
    sep = ""
  )
)

if (file.exists(cache_location)) {
  results <- readRDS(cache_location)$result
} else {
  results <- compute_SBC(
    prior_predictive_datasets,
    stan_backend,
    cache_mode = "results",
    cache_location = file.path(
      cache_dir,
      paste("results-n_iter_", n_sbc_iterations, "seed_", seed, sep = "")
    ),
    keep_fits = F,
    chunk_size = 4,
    thin_ranks = thin_by
  )
}
```

### Results

```{r}
#| label: analyse-results
#| echo: false
#| eval: false
ggplot(results$stats,
       aes(x = simulated_value, y = rank, colour = rhat)) +
  geom_point() +
  facet_wrap(vars(variable), scales = "free") +
  scale_colour_binned(
    type = "viridis",
    breaks = c(1.05),
    limits = c(1, 2.5),
    guide = guide_coloursteps(even.steps = TRUE, show.limits = TRUE)
  ) +
  theme(legend.position = "bottom")
```


```{r}
#| echo: false
#| eval: false
results$stats |>
  dplyr::select(sim_id, variable, simulated_value) |>
  tidyr::pivot_wider(names_from = variable, values_from = simulated_value) |>
  dplyr::left_join(
    results$stats |>
      dplyr::group_by(sim_id) |>
      dplyr::summarise(ess_bulk = min(ess_bulk), rhat = max(rhat)),
    by = c("sim_id")
  ) |>
  ggplot() +
  geom_point(aes(x = .panel_x, y = .panel_y, colour = rhat), alpha = .5) +
  ggforce::geom_autodensity(aes(group = rhat > 1.05, fill = ifelse(rhat > 1.05, 2, 1)),
                            alpha = .5,
                            position = "identity") +
  ggforce::facet_matrix(vars(unique(results$stats$variable)), layer.diag = 2) +
  scale_colour_binned(
    aesthetics = c("colour", "fill"),
    type = "viridis",
    breaks = c(1.05),
    limits = c(1, 2.5),
    guide = guide_coloursteps(
      even.steps = TRUE,
      show.limits = TRUE,
      title = "Rhat"
    )
  ) +
  theme(legend.position = "bottom")

```


A simple plot to compare the posterior samples and the ground truth values.

```{r}
#| label: prior-sbc_sim_estimated
#| warning: false
modified_plot_sim_estimated(
  results$stats[results$stats$max_rank == n_samples / thin_by - 1, ],
  variables = unique(results$stats$variable),
  alpha = 1,
  nrow = 3,
  facet_labeller = label_for_paper) +
  labs(x = "Ground truth", y = "Estimate")  +
  scale_x_continuous(breaks = scales::breaks_extended(3)) +
  scale_y_continuous(breaks = scales::breaks_extended(3))
```

The ECDF of the PIT values of the prior draw with regards to the posterior sample gives a principled way to check for the calibration of the model, as we can draw simultaneous 95% confidence intervals for the PIT-ECDFs.
The joint log-likelihood is a good test quantity for overall calibration of the inference, and shows here some possible calibration issues, although the number of SBC iterations is quite low.

```{r}
#| label: prior-sbc-pit-ecdf-diff
modified_ppc_pit_ecdf_grouped(
  pit = (results$stats$rank / results$stats$max_rank)[results$stats$max_rank == n_samples / thin_by - 1],
  group = results$stats$variable[results$stats$max_rank == n_samples / thin_by - 1],
  plot_diff = TRUE,
  nrow = 3,
  facet_labeller = label_for_paper) +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = scales::breaks_extended(3)) +
  scale_y_continuous(breaks = scales::breaks_extended(3))
```


